Splitting It Up: The \pkg{spduration} Split-Population Duration Regression Package for Time-varying Covariates

### 2017-07-11 R Journal/beger-hill-metternich-minhas-ward.tex
this version submitted to R Journal

\abstract{
We present an implementation of split-population duration regression in
the \CRANpkg{spduration} \citep{beger2016spduration} package for {\sf R} \citep{rstats2016} that allows for time-varying covariates. The statistical model accounts for units that are immune to a
certain outcome and not part of the duration process the researcher is
primarily interested in. We provide insights that if immune units exist,
we can significantly increase the predictive performance compared to
standard duration models. The package includes estimation and several
post-estimation methods for split-population Weibull and Loglogistic
models. We provide an empirical application to data on military
coups.
}

@@

The last component to complete the likelihood is the choice of a
distribution for the shape of the hazard rate. The \CRANpkg{spduration}
package implements two hazard rate shapes, Weibull and log-logistic:

\begin{eqnarray*}
\textrm{Weibull} \\
 f(t) & = & \alpha \lambda (\lambda t)^{\alpha - 1} e^{-(\lambda t)^\alpha} \\
  S(t) & = & e^{ -(\lambda t )^\alpha } \\
 h(t) & = & \alpha \lambda (\lambda t)^{\alpha-1} \\
\textrm{Log-logistic} \\
 f(t) & = & \frac{ \alpha \lambda (\lambda t)^{\alpha-1} }{ (1 + (\lambda t)^\alpha)^2 } \\
 S(t) & = & \frac{1}{ 1+  (\lambda t)^\alpha }  \\
 h(t) & = & \frac{ \alpha \lambda (\lambda t)^{\alpha-1} }{ 1+  (\lambda t)^\alpha }
\end{eqnarray*}

@@

\begin{eqnarray}
h(t, \pi) = \frac{f(t, \pi)}{S(t, \pi)} & = & \frac{ \pi(t) \times f(t) }{ (1-\pi(t)) + \pi(t) \times S(t) } \\
 \pi(t) & = & \frac{ 1-\pi }{ S(t) + (1-\pi) (1 - S(t)) }
\end{eqnarray}

@@

One useful extension of our project in the future would be to integrate and adapt the \code{Surv} class for survival data in \CRANpkg{survival} to the \CRANpkg{spduration} library. This would require some modification of the class as our model also requires information on the risk status of spells, but would give access to the much broader functionality, especially for descriptive summaries of data, in the \CRANpkg{survival} library.

Other currently available {\sf R} routines for estimation of split-population/cure rate duration models include the packages \CRANpkg{smcure}  \citep{cai2012smcureR}, described in \citet{cai2012smcure}, and \pkg{nltm} \citep{garibotti2010nltm}. \CRANpkg{smcure} implements semi-parametric proportional hazards and accelerated failure time cure models using estimation procedures presented in \citet{peng2003fitting} and \citet{zhang2007new}.\footnote{There is also the older \pkg{gfcure} package, which estimates the parametric accelerated failure time cure model discussed in \citet{peng1998generalized}. A version for Windows is available at \url{http://post.queensu.ca/~pengp/software.html}.} The package \pkg{nltm} will estimate a semi-parametric proportional hazard cure model, and a number of other survival models, using the approach developed in \citet{tsodikov2003semiparametric,tsodikov2007profile}. Semi-parametric estimation of the hazard function is attractive because it requires no assumption about its shape, and such assumptions can be difficult in practice to empirically evaluate. The main advantage \CRANpkg{spduration} offers is the ability to include time varying covariates, which gives it a broader range of practical applications, as \CRANpkg{smcure} and \pkg{nltm} will only accommodate case-based data. \CRANpkg{spduration} also features more post-estimation methods than alternative packages, including those discussed above to create and evaluate model predictions.

@@

\bibliography{beger-hill-metternich-minhas-ward}

\address{Andreas Beger\\
  Ward Associates\\
  Tallinn\\
  Estonia\\
  \email{adbeger@gmail.com}}

\address{Daniel W. Hill, Jr.\\
  Department of International Affairs\\
  School of Public and International Affairs, University of Georgia\\
  Candler Hall, Athens GA 30602\\
  United States of America\\
  \email{dwhill@uga.edu}}

>>>>

\abstract{
We present an implementation of split-population duration regression in
the \CRANpkg{spduration} \citep{beger2016spduration} package for {\sf R} that allows for time-varying covariates. The statistical model accounts for units that are immune to a
certain outcome and not part of the duration process the researcher is
primarily interested in. We provide insights that if immune units exist,
we can significantly increase the predictive performance compared to
standard duration models. The package includes estimation and several
post-estimation methods for split-population Weibull and Loglogistic
models. We provide an empirical application to data on military
coups.
}

@@

The last component to complete the likelihood is the choice of a
distribution for the shape of the hazard rate. The \pkg{spduration}
package implements two hazard rate shapes, Weibull and log-logistic:

\begin{eqnarray*}
\textrm{Weibull} \\
 f(t) & = & \alpha \lambda \left(\lambda t \right)^{\alpha - 1} e^{-\left(\lambda t \right)^\alpha} \\
  S(t) & = & e^{ -\left(\lambda t \right)^\alpha } \\
 h(t) & = & \alpha \lambda \left(\lambda t \right)^{\alpha-1} \\
\textrm{Log-logistic} \\
 f(t) & = & \frac{ \alpha \lambda \left(\lambda t \right)^{\alpha-1} }{ \left(1 + (\lambda t)^\alpha \right)^2 } \\
 S(t) & = & \frac{1}{ 1+  \left(\lambda t \right)^\alpha }  \\
 h(t) & = & \frac{ \alpha \lambda \left(\lambda t \right)^{\alpha-1} }{ 1+  \left(\lambda t \right)^\alpha }
\end{eqnarray*}

@@

\begin{eqnarray}
h \left(t, \pi \right) = \frac{f \left(t, \pi \right)}{S \left(t, \pi \right)} & = & \frac{ \pi \left(t \right) \times f \left(t \right) }{ \left(1-\pi \left(t \right) \right) + \pi \left(t \right) \times S \left(t \right) } \\
 \pi \left(t \right) & = & \frac{ 1-\pi }{ S \left(t \right) + \left(1-\pi \right) \left(1 - S \left(t \right) \right) }
\end{eqnarray}

@@

One useful extension of our project in the future would be to integrate and adapt the \code{Surv} class for survival data in \pkg{survival} to the \pkg{spduration} library. This would require some modification of the class as our model also requires information on the risk status of spells, but would give access to the much broader functionality, especially for descriptive summaries of data, in the \pkg{survival} library.

Other currently available {\sf R} routines for estimation of split-population/cure rate duration models include the packages \CRANpkg{smcure}  \citep{cai2012smcureR}, described in \citet{cai2012smcure}, and \pkg{nltm} \citep{garibotti2010nltm}. \pkg{smcure} implements semi-parametric proportional hazards and accelerated failure time cure models using estimation procedures presented in \citet{peng2003fitting} and \citet{zhang2007new}.\footnote{There is also the older \pkg{gfcure} package, which estimates the parametric accelerated failure time cure model discussed in \citet{peng1998generalized}. A version for Windows is available at \url{http://post.queensu.ca/~pengp/software.html}.} The package \pkg{nltm} will estimate a semi-parametric proportional hazard cure model, and a number of other survival models, using the approach developed in \citet{tsodikov2003semiparametric,tsodikov2007profile}. Semi-parametric estimation of the hazard function is attractive because it requires no assumption about its shape, and such assumptions can be difficult in practice to empirically evaluate. The main advantage \pkg{spduration} offers is the ability to include time varying covariates, which gives it a broader range of practical applications, as \pkg{smcure} and \pkg{nltm} will only accommodate case-based data. \pkg{spduration} also features more post-estimation methods than alternative packages, including those discussed above to create and evaluate model predictions.

@@

\bibliography{beger-hill-metternich-minhas-ward}
\newpage
\address{Andreas Beger\\
  Predictive Heuristics\\
  \email{adbeger@gmail.com}}

\address{Daniel W. Hill, Jr.\\
  Department of International Affairs\\
  University of Georgia\\
  Athens, GA\\
  United States of America\\
  \email{dwhill@uga.edu}}
